---
title: "Exploratory analysis"
author: "Jennah Gosciak"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
knit: (function(inputFile, encoding) { 
  rmarkdown::render(inputFile,
  encoding=encoding,
  output_file="01_analysis.html")})
output:
  html_document:
    toc: true
    theme: paper
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F,
                      fig.width = 6, fig.height = 10)
```

```{r, load packages}
# load packages
library(tidyverse)
library(magrittr)
library(readr)
library(dplyr)
library(ggplot2)
library(dbplyr)
library(DBI)
library(RPostgres)
library(assertr)
library(RColorBrewer)
library(sf)
library(fs)
library(purrr)
library(paletteer)
library(mapview)
library(leaflet)
library(sf)
library(gt)

root <- "C:/Users/jg6849/Documents/Github/SpatialAnalysis_Visualization/rproj_1"
inp <- path(root, "1_input")
out <- path(root, "2_output")
figs <- path(out, "figs")

source("99_common.R")
```
# Data source
Data is MapPLUTO, which is available from [NYC Open Data](https://data.cityofnewyork.us/City-Government/Primary-Land-Use-Tax-Lot-Output-Map-MapPLUTO-/f888-ni5f) and provided as spatial data from the [New York City Department of City Planning](https://www1.nyc.gov/assets/planning/download/zip/data-maps/open-data).

Other data sources include:
* Borough shapefiles from [DCP](https://www1.nyc.gov/site/planning/data-maps/open-data.page)
* NTA shapefiles from [DCP](https://www1.nyc.gov/site/planning/data-maps/open-data.page)
* Zoning amendments since 1977 from [DCP](https://www1.nyc.gov/site/planning/data-maps/open-data.page)

# Load PLUTO data
Load data for 2020 and 2010 (PLUTO)
```{r, load spatial data}
## load shapefiles of pluto data
temp <- tempfile()
temp2 <- tempfile()
download.file("https://www1.nyc.gov/assets/planning/download/zip/data-maps/open-data/nyc_mappluto_20v8_arc_shp.zip", temp)
unzip(zipfile = temp, exdir = temp2)
pluto_20shp <- read_sf(path(temp2, "MapPLUTO.shp")) %>%
  rename_all(~ str_to_lower(.)) %>%
  st_transform(2263) 
unlink(c(temp, temp2))
# 
temp <- tempfile()
temp2 <- tempfile()
download.file("https://www1.nyc.gov/assets/planning/download/zip/data-maps/open-data/mappluto_10v1.zip", temp)
unzip(zipfile = temp, exdir = temp2)

pluto_10shp <- map2_dfr(
  c(
    "Manhattan", "Brooklyn", "Queens", "Bronx",
    "Staten_Island"
  ),
  c("MN", "BK", "QN", "BX", "SI"),
  ~ read_sf(path(temp2, "MapPLUTO_10v1", .x, str_c(.y, "MapPLUTO.shp")))
)  %>%
  rename_all(~str_to_lower(.)) %>%
  st_transform(2263)
unlink(c(temp, temp2))
# 
pluto_join <- pluto_20shp %>% 
  st_centroid() %>%  
  st_transform(2263) %>% 
  st_join(pluto_10shp, join = st_within, suffix = c("_20", "_10")) %>% 
  mutate(join_result = case_when(is.na(bbl_10) ~ "Did not join",
                                 TRUE ~ "Joined")) %>% 
  select(starts_with("bbl_"), starts_with("unitsres"),
         starts_with("bldgclass"),
         starts_with("borough")) %>% 
  # need to drop one building bc of duplicates
  # owned by Con Ed
  # other building creates duplicates in 2020
  filter(bbl_10 != "2027810800",
         !(bbl_10 == "4008500050" & bbl_20 == "1018190100"))

pluto_join %>% 
  as.data.frame() %>% 
  eeptools::isid("bbl_20") %>% 
  stopifnot()

# this is less than 300
pluto_join %>% 
  as.data.frame() %>% 
  filter(is.na(bbl_10))

# drop them, small number in comparison to rest of the data
pluto_join <- pluto_join %>% 
  filter(!is.na(bbl_10))

# bbl 10 duplicates
# make sure to drop them when comparing 2010 and 2020
# otherwise there will be duplicates
pluto_join %>% 
  as.data.frame() %>% 
  filter(!is.na(bbl_10)) %>% 
  group_by(bbl_10) %>% 
  filter(n() > 1)

pluto_join %>% 
  as.data.frame() %>% 
  filter(!is.na(bbl_20)) %>% 
  group_by(bbl_20) %>% 
  filter(n() > 1)
```
```{r, check pt data}
# use point version for comparing old and new data
pluto_20pts <- pluto_20shp %>% 
  # better to use centroids
  # fewer cases aren't merging (only 217)
  st_centroid() %>% 
  st_transform(2263)

# check the sum of cases that aren't intersecting
pluto_check <- pluto_20pts %>% 
  st_within(pluto_20shp)
sum(lengths(pluto_check) == 0)
```

# Background analysis on rezoning
## Load geographic shapefiles
* Neighborhood (using NTAs)
* Zoning amendments, from NYC DCP
* Borough shapefiles, from NYC DCP
```{r, load rezoning areas}
url_zoning <- "https://services5.arcgis.com/GfwWNkhOj9bNBqoJ/arcgis/rest/services/nyzma/FeatureServer/0/query?where=1=1&outFields=*&outSR=4326&f=pgeojson"
url_boro <- "https://services5.arcgis.com/GfwWNkhOj9bNBqoJ/arcgis/rest/services/NYC_Borough_Boundary/FeatureServer/0/query?where=1=1&outFields=*&outSR=4326&f=pgeojson"
url_nhood <- "https://services5.arcgis.com/GfwWNkhOj9bNBqoJ/arcgis/rest/services/NYC_Neighborhood_Tabulation_Areas_2020/FeatureServer/0/query?where=1=1&outFields=*&outSR=4326&f=pgeojson"

rz <- st_read(url_zoning)
rz <- rz %>% 
  # convert from epoch time
  mutate(EFFECTIVE_DATE = as.POSIXct(EFFECTIVE/1000, origin = "1970-01-01"),
         EFFECTIVE_YEAR = lubridate::year(EFFECTIVE_DATE),
         EFFECTIVE_CAT = str_extract(cut(EFFECTIVE_YEAR, breaks = 7), "[0-9]+(?=])") %>% 
           if_else(is.na(.), "No year provided", .),
         EFEFCTIVE_CAT = str_c(as.numeric(EFFECTIVE_CAT) - 6, "-", EFFECTIVE_CAT),
         EFFECTIVE_CAT2 = str_extract(cut(EFFECTIVE_YEAR, seq(1976, 2020, 2)), "[0-9]+(?=])") %>% 
           if_else(is.na(.), "No year provided", .),
           EFFECTIVE_CAT2 = str_c(as.numeric(EFFECTIVE_CAT2) - 1, "-", EFFECTIVE_CAT2)) %>% 
    st_transform(2263)

# rz file for 2010 - 2020
rz_2010_2020 <- rz %>% 
  filter(EFFECTIVE_YEAR %in% c(2010:2020),
         !is.na(EFFECTIVE_YEAR),
         STATUS == "Adopted")

boro <- st_read(url_boro) %>% 
  st_transform(2263)

nhood <- st_read(url_nhood) %>% 
    st_transform(2263)
```
```{r}
# produce pluto file restricted to rezoned areas between 2010 and 2020
pluto_rz_int <- pluto_join %>% 
  st_intersects(rz_2010_2020)
pluto_rz <- pluto_join[lengths(pluto_rz_int) > 0, ]
pluto_rz %>%
  as.data.frame()

pluto_rz %>% 
  saveRDS(path(out, "pluto_rz.RDS"))
```

```{r, plot rezoning areas, fig.height = 6, fig.width = 10}
ggplot() +
  geom_sf(data = boro, fill = "lightgray", color = NA, alpha = 0.3) +
  geom_sf(data = nhood, color = "darkgray", alpha = 0.3, fill = NA, size = 0.1) +
  geom_sf(data = rz, fill = paletteer_d("rcartocolor::Earth", 6)[[6]],
          color = NA) +
  theme_classic() +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

ggsave(path(figs, "1_rezone_map.pdf"))
```

## Background statistisc on rezoned area
* Percent of total area citywide and by borough
* Number of rezonings by NTA

```{r, area ratios}
rz_adpt <- rz %>% 
  filter(STATUS == "Adopted")

## Square miles of rezoning
# all 
rz_adpt %>% 
  st_area(geometry) %>% 
  sum() %>% 
  as.numeric() %>% 
  `/`(5280^2)

# since 2000, with a nonmissing date
rz_adpt %>% 
  filter(EFFECTIVE_YEAR >= 2000 & !is.na(EFFECTIVE_YEAR)) %>% 
  st_area(geometry) %>% 
  sum() %>% 
  as.numeric() %>% 
  `/`(5280^2)

# missing date
rz_adpt %>% 
  filter(is.na(EFFECTIVE_YEAR)) %>% 
  st_area(geometry) %>% 
  sum() %>% 
  as.numeric() %>% 
  `/`(5280^2)

# share of total nyc area
rz_area <- rz_adpt %>% 
  st_transform(2263) %>% 
  st_area(geometry) %>% 
  sum() %>% 
  as.numeric()

rz_area_2010 <- rz_adpt %>% 
  filter(EFFECTIVE_YEAR %in% c(2010:2020)) %>% 
  st_area(geometry) %>% 
  sum() %>% 
  as.numeric()

rz_area_na <- rz_adpt %>% 
  filter(is.na(EFFECTIVE_YEAR)) %>% 
  st_area(geometry) %>% 
  sum() %>% 
  as.numeric()

boro_area <- boro %>% 
  st_area(geometry) %>% 
  sum() %>% 
  as.numeric()


rz_stats <- tibble(area_per = c((rz_area * 100 / boro_area),
                     (rz_area_2010 * 100 / boro_area)),
       file = c("All years",
                "2010-2020"),
       BoroName = rep("Citywide", 2))
rz_stats
```
* Share/percentage of rezoning area compared to borough area
```{r}
rz_boro <- rz_adpt %>% 
  st_join(boro, by = "intersects")

rz_boro_graph_df <- rz_boro %>% 
  mutate(file = "All years") %>% 
  bind_rows(filter(rz_boro, EFFECTIVE_YEAR %in% c(2010:2020)) %>% 
              mutate(file = "2010-2020")) %>% 
  mutate(rz_area_sqmi = as.numeric(st_area(geometry)) / (5280^2)) %>% 
  as.data.frame() %>% 
  group_by(file, BoroName) %>% 
  summarize(across(rz_area_sqmi, sum, na.rm = T)) %>% 
  full_join(boro, ., by = "BoroName") %>% 
  mutate(boro_area_sqmi = as.numeric(st_area(geometry)) / (5280^2)) %>% 
  as.data.frame() %>% 
  mutate(area_per = rz_area_sqmi * 100 / boro_area_sqmi) %>% 
  select(c("BoroName", "area_per", "file")) %>% 
  bind_rows(rz_stats) %>% 
  mutate(BoroName = factor(BoroName, levels = c("Manhattan", "Bronx",
                  "Brooklyn", "Queens", "Staten Island", "Citywide"),
         ordered = T)) %>% 
  mutate(file = factor(file, levels = c("All years", "2010-2020"), ordered = T))
rz_boro_graph_df
```

```{r, fig.width = 6, fig.height = 6}
rz_boro_graph_df  %>% 
  ggplot(aes(BoroName, area_per, fill = file)) +
  geom_col(position = position_dodge(0.9),
           width = 0.8) +
  geom_text(aes(label = round(area_per), group = file),
            position = position_dodge(0.9), vjust = -1) +
  scale_fill_paletteer_d("rcartocolor::Earth", name = "") +
  theme_classic() +
  labs(x = "Borough",
       y = "Percent of borough area")

ggsave(path(figs, "3_rezoning_stats.pdf"))
ggsave(path(figs, "3_rezoning_stats.png"))
```


```{r}
rz_stats %>% 
  ggplot(aes(reorder(file, -area_per),
               area_per)) +
  geom_col(fill = paletteer_d("rcartocolor::Earth", 2)[[2]]) +
  geom_text(aes(label = round(area_per)), vjust = -1) +
  theme_classic() +
  labs(x = "", y = "Percent of NYC area") +
  scale_y_continuous(limits = c(0, 60))

ggsave(path(figs, "0_rz_stats.pdf"), width = 5, height = 6)
ggsave(path(figs, "0_rz_stats.png"), width = 5, height = 6)
```
```{r, plot rezoning areas yr, fig.height = 6, fig.width = 10}
ggplot() +
  geom_sf(data = boro, fill = "lightgray", color = NA, alpha = 0.3) +
  geom_sf(data = nhood, color = "darkgray", alpha = 0.3, fill = NA, size = 0.1) +
  geom_sf(data = rz_adpt, aes(fill = EFFECTIVE_CAT),
          color = NA) +
  scale_fill_manual(values = c(paletteer_d("rcartocolor::Earth", 7), "dark gray"),
                    name = "Year adopted") +
  theme_classic() +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  labs(title = "Adopted rezonings by year adopted")
ggsave(path(figs, "2_rezoning_by_year.pdf"))
ggsave(path(figs, "2_rezoning_by_year.png"))
```
```{r, fig.height = 6, fig.width = 10}
filter(rz, STATUS == "Adopted" 
                        & EFFECTIVE_YEAR %in% c(2010:2020)) %>% 
  as.data.frame() %>% 
  mutate(area = st_area(geometry)) %>% 
  arrange(-area) %>% 
  head(10)
 
ggplot() +
  geom_sf(data = boro, fill = "lightgray", color = NA, alpha = 0.3) +
  geom_sf(data = nhood, color = "darkgray", alpha = 0.3, fill = NA, size = 0.1) +
  geom_sf(data = filter(rz_adpt, EFFECTIVE_YEAR %in% c(2010:2020)),
          aes(fill = EFFECTIVE_CAT2),
          color = NA) +
  scale_fill_manual(values = c(paletteer_d("rcartocolor::Earth", 7), "dark gray"),
                    name = "Year adopted") +
  theme_classic() +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  labs(title = "Adopted rezonings by year adopted")
ggsave(path(figs, "2_rezoning_by_year_2010_2020.pdf"))
ggsave(path(figs, "2_rezoning_by_year_2010_2020.png"))
```
```{r, fig.width = 10}
rz_nta_graph_df <- rz_adpt %>% 
  filter(EFFECTIVE_YEAR %in% c(2010:2020)) %>%
  mutate(geometry = st_centroid(geometry)) %>% 
  st_join(nhood, by = "intersects") %>% 
  as.data.frame() %>% 
  group_by(BoroName, NTAName) %>% 
  summarize(n_rz = n()) %>%
  arrange(-n_rz) %>% 
  mutate(Boro_NTA = str_wrap(NTAName, 10))

rz_nta_graph_df %>% 
  head(10) %>% 
  ggplot(aes(reorder(Boro_NTA, -n_rz), n_rz)) +
  geom_col(fill = paletteer_d("rcartocolor::Earth", 2)[[2]]) +
  geom_text(aes(label = n_rz), vjust = -1) +
  theme_classic() +
  labs(x = "Neighborhood",
       y = "Number of adopted rezonings") +
  scale_y_continuous(limits = c(0, 9), breaks = seq(0, 8, 2))

ggsave(path(figs, "4_count_rezonings_nta.pdf"))
ggsave(path(figs, "4_count_rezonings_nta.png"))
```
```{r}
pal <- paletteer_d("rcartocolor::Earth", 7)
rz_nta_graph_df2 <- rz_nta_graph_df %>% 
  arrange(-n_rz) %>% 
  left_join(nhood, ., by = "NTAName") %>% 
  filter(!(is.na(n_rz) | n_rz == 0))

  ggplot() +
  geom_sf(data = nhood, fill = "lightgray", color = NA,
          alpha = 0.3) +
  geom_sf(data = nhood, alpha = 0.3, fill = NA, 
          color = "darkgray", size = 0.1) +
  geom_sf(data = rz_nta_graph_df2, aes(fill = n_rz), color = NA) +
  scale_fill_gradient(name = "Number of rezonings",
                      low = "#F0F8FF", high = pal[7]) +
  theme_classic() +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

ggsave(path(figs, "5_num_rz_2010_2020.pdf"))
ggsave(path(figs, "5_num_rz_2010_2020.png"))
```


## Comparison of adopted vs. certified
```{r}
ggplot() +
  geom_sf(data = boro, fill = "lightgray", color = NA, alpha = 0.5) +
  geom_sf(data = nhood, color = "darkgray", alpha = 0.7, fill = NA, size = 0.1) +
  geom_sf(data = rz, aes(fill = STATUS),
          color = NA) +
  theme_classic() +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  scale_fill_manual(values = c(paletteer_d("rcartocolor::Earth", 6)[[6]], paletteer_d("rcartocolor::Earth", 2)[[2]]))
```
# Building code analysis
Recode building class values
```{r, compare changes}
# load html codebook
url <- "https://www1.nyc.gov/assets/finance/jump/hlpbldgcode.html"
othcodes <- tibble(
  `Building Code` = c("Q0", "RC", "RX", "RD", "RM", "RI", "RZ", "QG"),
  Description = c(
    "Park", "Mixture of commercial types",
    "Mixture of commerical, residential, and industrial",
    "Mixture residential", "Mixture commercial and residential",
    "Mixture commercial and industrial",
    "Mixture residential and industrial",
    "Community garden"
  )
)

bldgcodes <- htmltab::htmltab(doc = url) %>%
  rbind(othcodes)
```
```{r}
# use function to compare building classes
compare_bldgclass <- function(df) {
  df %>%
  as.data.frame() %>% 
  filter(bldgclass_20 != bldgclass_10 
         & !is.na(bldgclass_20) 
         & !is.na(bldgclass_10)) %>% 
  mutate(across(c("bldgclass_20", "bldgclass_10"), .fns = lst(orig = ~.))) %>%
  mutate(across(
    c("bldgclass_20", "bldgclass_10"),
    ~ (bldgcodes$Description[match(., bldgcodes$`Building Code`)])
  )) %>%
  verify((is.na(bldgclass_20_orig) | bldgclass_20_orig == "  ") == is.na(bldgclass_20)) %>%
  mutate(bldgclass_changes = str_c(bldgclass_10, " TO ", bldgclass_20)) %>% 
  # in order to avoid counting duplicate bbl_10 units
  group_by(bbl_10) %>% 
  mutate(unitsres_10 = case_when((n() > 1 & row_number() != 1) ~ NA_integer_,
                                 TRUE ~ unitsres_10)) %>% 
    ungroup() 
}

# update categories from codes to description
pluto_bldg <- pluto_join %>% 
  compare_bldgclass()

# same process, but with rezoned lots
pluto_bldg_rz <- pluto_rz %>% 
  compare_bldgclass()

pluto_bldg_desc %>%
  as.data.frame() %>% 
  group_by(bldgclass_changes) %>%
  # produce frequency and percent
  summarize(freq = n()) %>%
  ungroup() %>%
  mutate(per = round(freq * 100 / sum(freq, na.rm = T), 2)) %>%
  arrange(-per)

# only rezoned buildings
pluto_bldg_rz %>%
  as.data.frame() %>% 
  group_by(bldgclass_changes) %>%
  # produce frequency and percent
  summarize(freq = n()) %>%
  ungroup() %>%
  mutate(per = round(freq * 100 / sum(freq, na.rm = T), 2)) %>%
  arrange(-per)

# check units in cases with dupl bbl_10 values
pluto_bldg_rz %>% 
  group_by(bbl_10) %>% 
  filter(n() > 1) %>% 
  arrange(bbl_10) %>% 
  select(c("unitsres_10", "unitsres_20"), everything())
```

* Mostly small changes (1-family to 2-family and vice versa)
```{r, fig.width = 10}
pluto_bldg_desc %>%
  as.data.frame() %>% 
  group_by(bldgclass_changes) %>%
  # produce frequency and percent
  summarize(freq = n()) %>%
  ungroup() %>%
  mutate(per = round(freq * 100 / sum(freq, na.rm = T), 2)) %>%
  filter(per > 1) %>%
  arrange(-per) %>%
  head(5) %>% 
  # format text
  mutate(bldgclass_changes = str_wrap(str_to_sentence(bldgclass_changes), width = 15)) %>%
  # produce column graph
  ggplot(aes(reorder(bldgclass_changes, -per), per)) +
  geom_col(fill = paletteer_d("rcartocolor::Earth", 1)) +
  geom_text(aes(label = round(per, 1)), vjust = -1) +
  theme_classic() +
  labs(
    y = "Percentage of all BBLs with \n different building codes between 2010 and 2020",
    x = ""
  )

ggsave(path(figs, "6_bbl_bld_diff.pdf"))
ggsave(path(figs, "6_bbl_bld_diff.png"))
```
```{r}
# graph of average difference in units, and avg # of units in 2010 and 2020
pluto_bldg_desc %>% 
  as.data.frame() %>% 
  mutate(units_diff = unitsres_20 - unitsres_10) %>% 
  group_by(borough_20) %>% 
  summarize(across(c("units_diff", "unitsres_20", "unitsres_10"), mean, na.rm = T)) %>% 
  pivot_longer(-"borough_20") %>% 
  mutate(type = case_when(name == "units_diff" ~ "Difference in units",
                          name == "unitsres_20" ~ "Units in 2020",
                          name == "unitsres_10" ~ "Units in 2010")) %>% 
  ggplot(aes(borough_20, value)) +
  geom_col(aes(fill = type), width = 0.8, position = position_dodge(0.9)) +
  geom_text(aes(label = ifelse(round(value, 1) > 0, round(value, 1), NA_real_), group = type), position = position_dodge(0.9), vjust = -1, size = 3) + 
  scale_y_continuous(limits = c(0, 20)) +
  scale_fill_paletteer_d("rcartocolor::Earth", 1,
                         name = "") +
  theme_classic() +
  labs(x = "Borough",
       y = "Average number of units",
       title = "Buildings that experienced a change in building class between 2020 and 2010")

ggsave(path(figs, "7_unit_change.pdf"))
ggsave(path(figs, "7_unit_change.png"))
```
* Mostly small changes (1-family to 2-family and vice versa)
```{r, fig.width = 10}
pluto_bldg_rz %>%
  as.data.frame() %>% 
  group_by(bldgclass_changes) %>%
  # produce frequency and percent
  summarize(freq = n()) %>%
  ungroup() %>%
  mutate(per = round(freq * 100 / sum(freq, na.rm = T), 2)) %>%
  filter(per > 1) %>%
  arrange(-per) %>%
  head(5) %>% 
  # format text
  mutate(bldgclass_changes = str_wrap(str_to_sentence(bldgclass_changes), width = 15)) %>%
  # produce column graph
  ggplot(aes(reorder(bldgclass_changes, -per), per)) +
  geom_col(fill = paletteer_d("rcartocolor::Earth", 1)) +
  geom_text(aes(label = round(per, 1)), vjust = -1) +
  theme_classic() +
  labs(
    y = "Percentage of all BBLs with \n different building codes between 2010 and 2020 - rezoned areas",
    x = ""
  )

ggsave(path(figs, "8_bbl_bld_diff_rz.pdf"))
ggsave(path(figs, "8_bbl_bld_diff_rz.png"))
```
```{r}
# graph of average difference in units, and avg # of units in 2010 and 2020
pluto_bldg_rz %>% 
  as.data.frame() %>% 
  mutate(units_diff = unitsres_20 - unitsres_10) %>% 
  group_by(borough_20) %>% 
  summarize(across(c("units_diff", "unitsres_20", "unitsres_10"), mean, na.rm = T)) %>% 
  pivot_longer(-"borough_20") %>% 
  mutate(type = case_when(name == "units_diff" ~ "Difference in units",
                          name == "unitsres_20" ~ "Units in 2020",
                          name == "unitsres_10" ~ "Units in 2010")) %>% 
  ggplot(aes(borough_20, value)) +
  geom_col(aes(fill = type), width = 0.8, position = position_dodge(0.9)) +
  geom_text(aes(label = ifelse(round(value, 1) > 0, round(value, 1), NA_real_), group = type), position = position_dodge(0.9), vjust = -1, size = 3) + 
  scale_y_continuous(limits = c(0, 20)) +
  scale_fill_paletteer_d("rcartocolor::Earth", 1,
                         name = "") +
  theme_classic() +
  labs(x = "Borough",
       y = "Average number of units",
       title = "Buildings that experienced a change in building class between 2020 and 2010 - rezoned areas")

ggsave(path(figs, "9_unit_change_rz.pdf"))
ggsave(path(figs, "9_unit_change_rz.png"))
```

# Same building class analysis, coarser categories
* Only using lots in rezoned areas
```{r, message = T, warning = T}
varlist_recode <- readxl::read_excel(path(inp, "bldcodes_recode.xlsx")) %>% 
  separate(Character, c(str_c("char", 1:4)), sep = ",") %>% 
  separate(Codes, c(str_c("code", 1:10)), sep = ",") %>% 
  pivot_longer(starts_with("char"), values_to = "bldgclass_20_char") %>% 
  select(-"name") %>% 
  pivot_longer(starts_with("code"), values_to = "bldgclass_20") %>% 
  select(-"name") %>% 
  mutate(across(c("bldgclass_20_char",
                  "bldgclass_20"), ~str_trim(., "both")))
varlist_recode

varlist_char <- unique(select(varlist_recode, c("Aggregated category", 
                                            "bldgclass_20_char"))) %>% 
  filter(!is.na(bldgclass_20_char))

varlist_code <- unique(select(varlist_recode, c("Aggregated category",
                                                "bldgclass_20"))) %>% 
  filter(!is.na(bldgclass_20))

stopifnot(eeptools::isid(varlist_char, "bldgclass_20_char"))
stopifnot(eeptools::isid(varlist_code, "bldgclass_20"))
```
```{r}
pluto_use_change <- pluto_rz %>% 
  as.data.frame() %>%
  mutate(bldgclass_20_char = str_extract(bldgclass_20, "[:alpha:]"),
         bldgclass_10_char = str_extract(bldgclass_10, "[:alpha:]")) %>%
  select(c("geometry", "bbl_20", "bbl_10", "bldgclass_20_char", "bldgclass_20",
           "bldgclass_10_char", "bldgclass_10", "borough_20")) %>% 
  left_join(varlist_char, by = "bldgclass_20_char") %>% 
  left_join(varlist_code, by = "bldgclass_20", suffix = c("_char", "_code"))  %>% 
  mutate(`Aggregated category 20` = case_when(!is.na(`Aggregated category_code`) ~ `Aggregated category_code`,
                                           TRUE ~ `Aggregated category_char`)) %>% 
  select(-c("Aggregated category_char", "Aggregated category_code")) %>% 
  left_join(rename(varlist_char, bldgclass_10_char = bldgclass_20_char),
            by = "bldgclass_10_char") %>% 
  left_join(rename(varlist_code, bldgclass_10 = bldgclass_20),
            by = "bldgclass_10", suffix = c("_char", "_code")) %>% 
  mutate(`Aggregated category 10` = case_when(!is.na(`Aggregated category_code`) ~ `Aggregated category_code`,
                                           TRUE ~ `Aggregated category_char`)) %>% 
  select(-c("Aggregated category_char", "Aggregated category_code"))

pluto_use_change %>% 
  filter(!is.na(bldgclass_20), is.na(`Aggregated category 20`)) %>% 
  nrow() %>% 
  equals(0) %>% 
  stopifnot()
```
```{r, fig.width = 10}
use_change_graph_df <- pluto_use_change %>%
  filter(`Aggregated category 20` != `Aggregated category 10`) %>%
  mutate(change_cat = str_c(`Aggregated category 10`, " to ", `Aggregated category 20`)) %>%
  group_by(change_cat) %>%
  summarize(freq = n()) %>%
  ungroup() %>%
  mutate(per = round(freq * 100 / sum(freq, na.rm = T), 1)) %>%
  # format text
  mutate(change_cat = str_wrap(str_to_sentence(change_cat), width = 15)) %>%
  arrange(-per)
```

```{r}
gt1 <- use_change_graph_df %>% 
  gt() %>% 
  gt_style() %>% 
  cols_label(change_cat = "",
             freq = "Frequency",
             per = "Percent")
gt1
gtsave(gt1, path(figs, "dist_changes.html"))
```

```{r}
use_change_graph_df  %>%
  filter(per >= 3) %>%
  ggplot(aes(reorder(change_cat, -per), per)) +
  geom_col(fill = paletteer_d("rcartocolor::Earth", 2)[[2]]) +
  geom_text(aes(label = per), vjust = -1) +
  theme_classic() +
  labs(
    y = "Percentage of all BBLs with \n different building codes between 2010 and 2020",
    x = ""
  )

ggsave(filename = path(figs, "10_rz_bld_change_coarse.pdf"), width = 10)
ggsave(filename = path(figs, "10_rz_bld_change_coarse.png"), width = 10)
```
```{r, check strange obs}
pluto_use_change  %>% 
  filter(`Aggregated category 20` != `Aggregated category 10`) %>% 
  mutate(change_cat = str_to_sentence(str_c(`Aggregated category 10`, " to ", `Aggregated category 20`))) %>% 
  filter(change_cat == "Less than six families to vacant")

pluto_use_change  %>% 
  filter(`Aggregated category 20` != `Aggregated category 10`) %>% 
  mutate(change_cat = str_to_sentence(str_c(`Aggregated category 10`, " to ", `Aggregated category 20`))) %>% 
  filter(change_cat == "Less than six families to vacant") %>% 
  mutate(diff_bbl = if_else(bbl_20 != bbl_10, 1, 0)) %>% 
  pull(diff_bbl) %>% 
  mean(na.rm = T) %>% 
  `*`(100) %>%
  round() %>% 
  str_c("% of records have different bbl_20 and bbl_10 values")

pluto_use_change  %>% 
  filter(`Aggregated category 20` != `Aggregated category 10`) %>% 
  mutate(change_cat = str_to_sentence(str_c(`Aggregated category 10`, " to ", `Aggregated category 20`))) %>% 
  filter(change_cat == "Less than six families to vacant") %>% 
  mutate(diff_bbl = if_else(bbl_20 != bbl_10, 1, 0)) %>% 
  pull(diff_bbl) %>% 
  mean(na.rm = T) %>% 
  `*`(100) %>%
  round() %>% 
  str_c("% of records have different bbl_20 and bbl_10 values")
```
```{r, graph count}
use_change_graph_df %>%
  head(10) %>% 
  ggplot(aes(reorder(change_cat, -freq), freq)) +
  geom_col(fill = paletteer_d("rcartocolor::Earth", 2)[[2]]) +
  geom_text(aes(label = freq), vjust = -1) +
  theme_classic() +
  scale_y_continuous(limits = c(0, 800)) +
  labs(
    y = "Number of 2020 BBLs with \n different building codes between 2010 and 2020",
    x = ""
  )
```

```{r}
use_change_boro <- pluto_use_change %>%
  filter(`Aggregated category 20` != `Aggregated category 10`) %>%
  mutate(change_cat = str_c(`Aggregated category 10`, " to ", `Aggregated category 20`)) %>%
  group_by(change_cat, borough_20) %>%
  summarize(freq = n()) %>%
  group_by(borough_20) %>%
  mutate(per = round(freq * 100 / sum(freq, na.rm = T), 1)) %>%
  # format text
  mutate(change_cat = str_wrap(str_to_sentence(change_cat), width = 15)) %>%
  arrange(-per) %>%
  filter(per >= 3)

use_change_boro %>% 
  ggplot(aes(reorder(change_cat, -per), per)) +
  geom_col(aes(fill = borough_20)) +
  geom_text(aes(label = per), vjust = -1) +
  scale_fill_paletteer_d("rcartocolor::Earth") +
  theme_classic() +
  labs(
    y = "Percentage of all BBLs with \n different building codes between 2010 and 2020",
    x = ""
  )
```

